---
title: "Configuring categorical functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Configuring categorical functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(scicalc)
```

Several categorical functions ship with standard mappings that you can override
through options: `racen()` maps race strings to codes, while `agec()` and
`bmic()` bin a number into codes using fixed bands.

## Race codes (`racen()`)

`racen()` maps race strings to numeric codes. There are six built-in categories:

```{r}
racen(c(
  "WHITE",
  "BLACK",
  "ASIAN",
  "AMERICAN INDIAN OR ALASKA NATIVE",
  "PACIFIC ISLANDER",
  "OTHER"
))
```

Set `scicalc.racen_config` (a named numeric vector, names case-insensitive) to change them.

## Override a built-in code

```{r}
options(scicalc.racen_config = c(white = 10, black = 20))
racen(c(
  "WHITE",
  "BLACK",
  "ASIAN",
  "AMERICAN INDIAN OR ALASKA NATIVE",
  "PACIFIC ISLANDER",
  "OTHER"
))
```

`WHITE` and `BLACK` use the overrides; all others keeps their default.

## Add a new category

Names that aren't built-ins become new mappings, matched before the defaults.

```{r}
options(scicalc.racen_config = c("multiple" = 7))
racen(c("WHITE", "MULTIPLE"))
```

## Age and BMI bands (`agec()`, `bmic()`)

`agec()` and `bmic()` bin a number into codes using fixed bands (FDA age bands,
WHO BMI bands). By default:

```{r}
agec(c(1, 10, 30, 70))
```

Set `scicalc.agec_config` to a data frame of bands — `label`, `min` (lower
bound), and `code` — to replace them. Bands are half-open `[min, next min)`,
with the top band open-ended.

```{r}
options(scicalc.agec_config = data.frame(
  label = c("child", "adult", "senior"),
  min   = c(0, 18, 65),
  code  = c(1, 2, 3)
))
agec(c(5, 30, 70))
```

`bmic()` works the same way through `scicalc.bmic_config`:

```{r}
options(scicalc.bmic_config = data.frame(
  label = c("lean", "high"),
  min   = c(0, 27),
  code  = c(1, 2)
))
bmic(c(22, 30), age = c(40, 40))
```

With the options set, the package startup message reports the active
configuration:

```{r, echo = FALSE, comment = ""}
scicalc:::scicalc_options_message()
```

```{r, include = FALSE}
options(
  scicalc.racen_config = NULL,
  scicalc.agec_config = NULL,
  scicalc.bmic_config = NULL
)
```
